Data Management and Domain Adjusting Strategies for Implementing Parallel SPH Method in JPUE Simulation

### 2016-01-29 Paper for submission.tex
The final manuscript
01292016

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
This paper presents a parallel implementation of smoothed particle hydrodynamics (SPH) method using the message passing interface (MPI) standard to simulate a turbulent jet or plume which is ejected from a nozzle into a uniform environment(JPUE). Background grid is used to reduce neighbors searching cost and to decompose (redecompose) the domain. Space Filing Curve(SFC) based index is adopted to assign an unique identifier to each background grid. As simulation of JUPE requires adding of new particles during simulation, time-dependent SFC based indexes are assigned to particles to guarantee uniqueness of identifiers.  
%to distinguish particles added at the same position but different simulation time step. 
Both particle and background grid are managed by hash tables which ensure quick and flexible accessing, adding and deleting of element. %Additional link list attached to the hashtable is used to handle hash conflicts.
A SFC based three dimensional domain decomposition and a dynamic load balancing strategy are implemented to ensure good load balance. In addition, computational domain is adjusted during simulation to reduce computational cost. 
Numerical tests show that our code has acceptable strong scalability and good weak scalability.
These strategies can be further applied to many other implementations of meshfree methods, especially those implementations that require irregular
particle distributions, adding and deleting of particles and leverage adaptive background grids.
\end{abstract}

@@

There are several review papers \cite{monaghan1992smoothed, monaghan2005smoothed, price2012smoothed, monaghan2012smoothed} which provide a pretty comprehensive descriptions of SPH. These may be consulted for additional detail.%We will not cover the basic theory of SPH in this paper.

@@

Experiments have been carried out on the computational cluster of Center for Computational Research (CCR) at Buffalo. Two types of processors -- 8 core L5520/L5630 running at 2.27/2.13GHz clock rate with 3GB memory per core on a Mellanox Infiniband network and 12 core E5645 running at 2.40GHz clock rate with 4GB memory per core on a Q-Logic Infiniband network. Each node comprised of two sockets with several of these processors. Memory and level 3 cache are shared on each node.
An important caveat in interpreting these results is that they were obtained in settings where the processors were obtained under normal scheduling (hence readily available) wherein the nodes may be shared with other users. Thus our results may suffer from contention for memory, CPU-memory bandwidth and network access. As of this writing, we are waiting for exclusive access which will undoubtedly provide better results. On the other hand, these results are representative of what will be available during production use.
The initial domain is $[-4.8km,4.8km]\times [-4.8km,4.8km] \times [0km, 6km]$, with smoothing length (we set initial intervals between particles equal to smoothing length) equals to 200m and 100m respectively for test case1 and test case2, respectively. The computational work load of test case 2 is 8 times of that of the test case 1. The simulations run for 20s physical time.  Wall clock time and speed up are shown in figure \ref{fig:2cases_time} and \ref{fig:2cases_efficiency}. Linear speed up is observed when number of processes is smaller than 16. Test case2 shows better speed up than test case1 which implies that the overhead of strong scalability can be increased by increasing total amount of work load. We also compared the performance of 2D domain decomposition and 3D domain decomposition. As shown by figure \ref{fig:2D_vs_3D_efficiency}, 2D domain decomposition shows a little bit better speed up than 3D domain decomposition for 32 processes. This can be explained by the fact that 2D domain decomposition can get more regular subdomain and as a results, need less communication. When number of processes increase, it becomes harder for 2D decomposition to balance workload, that is why 3D domain decomposition has better speed up for processes number equals to 64 and 128. The weak scalability test is conducted with the same initial domain and various smoothing length. Each simulation runs for 400 loops. The average number of real particles of each process keep constant at $25900$. As shown in figure \ref{fig:weak_loglog}, simulation time are almost constant with some minor fluctutations around the average value. These fluctuations are due to sharing nodes with other users. In addition, difference in simulation time for different processors are obvious.\\

@@

\subfloat[A plume from Sakurajima Volcano eruption\cite{PlumePhoto}]{\includegraphics[width=3.8cm,height=4.4cm]{plume_photo}}

@@

\subfloat[Simulation results: mass fraction of erupted material at 290 seconds]{\includegraphics[width=3.8cm,height=4.4cm]{Plume_simulation}}

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
This paper presents a parallel implementation of smoothed particle hydrodynamics (SPH) method using the message passing interface (MPI) standard to simulate a turbulent jet or plume which is ejected from a nozzle into a uniform environment(JPUE). Background grid is used to reduce neighbors searching cost and to decompose (redecompose) the domain. Space Filing Curve(SFC) based index is adopted to assign an unique identifier to each background grid. As simulation of JUPE requires adding of new particles during simulation, time-dependent SFC based indexes are assigned to particles to guarantee uniqueness of identifiers.  
Both particle and background grid are managed by hash tables which ensure quick and flexible accessing, adding and deleting of element.
A SFC based three dimensional domain decomposition and a dynamic load balancing strategy are implemented to ensure good load balance. In addition, computational domain is adjusted during simulation to reduce computational cost. 
Numerical tests show that our code has acceptable strong scalability and good weak scalability.
These strategies can be further applied to many other implementations of meshfree methods, especially those implementations that require irregular
particle distributions, adding and deleting of particles and leverage adaptive background grids.
\end{abstract}

@@

There are several review papers \cite{monaghan1992smoothed, monaghan2005smoothed, price2012smoothed, monaghan2012smoothed} which provide a pretty comprehensive descriptions of SPH. These may be consulted for additional details. %We will not cover the basic theory of SPH in this paper.

@@

Experiments have been carried out on the computational cluster of Center for Computational Research (CCR) at Buffalo. 
Three types of processors -- 8 core L5520/L5630 running at 2.27/2.13GHz clock rate with 3GB memory per core on a Mellanox Infiniband network and 12 core E5645 running at 2.40GHz clock rate with 4GB memory per core on a Q-Logic Infiniband network. Each node comprise of two sockets with several of these processors. Memory and level 3 cache are shared on each node. 
The initial domain is $[-4.8km,4.8km]\times [-4.8km,4.8km] \times [0km, 6km]$, with smoothing length (we set initial intervals between particles equal to smoothing length) equals to 200m and 100m respectively for test case1 and test case2. The computational work load of test case 2 is 8 times of that of the test case 1. The simulations run for 20s physical time.  Wall clock time and speed up are shown in figure \ref{fig:2cases_time} and \ref{fig:2cases_efficiency}. Linear speed up is observed when number of processes is smaller than 16. Test case2 shows better speed up than test case1 which implies that the overhead of strong scalability can be increased by increasing total amount of work load. We also compared the performance of 2D domain decomposition and 3D domain decomposition. As shown by figure \ref{fig:2D_vs_3D_efficiency}, 2D domain decomposition shows a little bit better speed up than 3D domain decomposition for 32 processes. This can be explained by the fact that 2D domain decomposition can get more regular subdomain and as a results, need less communication. When number of processes increase, it becomes harder for 2D decomposition to balance workload, that is why 3D domain decomposition has better speed up for processes number equals to 64 and 128.\\ 
The weak scalability test is conducted with the same initial domain and various smoothing length. Each simulation runs for 400 loops. The average number of real particles of each process keeps constant at $25900$. 
Tests are carried out under two different settings: the normal scheduling (hence readily available) wherein the nodes may be shared with other users and exclusive access setting. Results of normal setting tests may suffer from contention for memory, CPU-memory bandwidth and network access but are representative of what will be available during production use.
As shown in figure \ref{fig:weak_loglog}, simulation time are almost constant with some minor fluctutations except for an abrupt jump from 132 processors to 620 processors. This jump, based on monitor data from nodes health check system on CCR, is due to a daemon which eats up a cpu. This inteference happens for all tests whose number of processes larger than or equal to 620. The wall clock time of exclusive test is slightly smaller than non-exclusive test which consistent with our anticipation. 
%These fluctuations are due to sharing nodes with other users. In addition, difference in simulation time for different processors are obvious.

@@

\subfloat[A plume from Sakurajima Volcano eruption\cite{PlumePhoto}]{\includegraphics[width=3.8cm,height=4.0cm]{plume_photo}}

@@

\subfloat[Simulation results: mass fraction of erupted material at 220 seconds]{\includegraphics[width=3.8cm,height=4.0cm]{Plume_simulation}}
